Predictions From Day 0 Community Structures

### 2016-08-10 code/HumanCdGF_microbiota_predict.Rmd
Use RF pedictors to create plots

```{r, message=FALSE}
if(!require(randomForest)) install.packages('randomForest', repos = 'http://cran.us.r-project.org')
library(randomForest)
if(!require(miscTools)) install.packages('miscTools', repos = 'http://cran.us.r-project.org')
library(miscTools)
if(!require(ggplot2)) install.packages('ggplot2', repos = 'http://cran.us.r-project.org')
library(ggplot2)
```

@@

for (i in 1:10){
  predict_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
  rsq_df[i] <- RF$rsq[length(RF$rsq)]
  # sort the features by their effect on % increase in the standard error
  importance_df[ , i] <- importance(RF)[,1]
  colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
# fit the model back on the datar and tack it to the end of the counts file
  random_forest_output <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day', 'log_cfu') ],
                              as.data.frame(predict(RF, rel_abund_stool)), 
                              by='Mouse_ID', by.y='row.names')
  colnames(random_forest_output)[3:4] <- c('observed', 'predicted')
  prediction_df <- rbind(prediction_df, random_forest_output)  
}

ggplot(data = prediction_df, aes(x=observed, y=predicted, color = day)) + geom_point()

rsq_df <- t(rsq_df)
rsq_df <- as.data.frame(cbind(rsq_df, day = c(1:10)))
ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_bar(stat = 'identity')

#create function for random forest
run_rf_on <- function(DATAFRAME, RESPONSE,  OTUs = OTU_list){
set.seed(2)
     #
     Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
    RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = train, importance = TRUE)
     R2 <- rSquared(test[,RESPONSE], test[,RESPONSE] - predict(RF,test))
     MSE <- mean((test[,RESPONSE] - predict(RF,test))^2)
     IMPORT <- importance(RF)
     FOREST <- RF
     PLOT<- ggplot(data = data.frame(actual=test[,RESPONSE],pred=predict(RF,test)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
     ##MEAN_DECREASE <- varImpPlot(RF)
     return(list(rf=FOREST,r2=R2,mse=MSE,import=IMPORT,plot=PLOT))
}

# let's plot the top 12 features
#n_features <- 12
#importance_subset <- importance_sorted[1:n_features]

# let's get Rsq for the subset
#rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
                           importance=TRUE, ntree=n_trees)
# write(paste("partial", n_features, rf_partial$rsq[n_trees], "\t"), file="data/process/random_forest.data", append=T)

@@

```{r}
Day_Euth_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_Euth_cdiff_cfu')
Day_Euth_cdiff_cfu_rf$plot
```

% Variance explained `r Day_Euth_cdiff_cfu_rf$rf$rsq[length(Day_Euth_cdiff_cfu_rf$rf$rsq)]*100`  
r^2^ = `r Day_Euth_cdiff_cfu_rf$r2`  
MSE = `r Day_Euth_cdiff_cfu_rf$mse`

@@

##Can we predict moribundity with the community structure at Day 0?
```{r}
##Fix to be categorical, and not regression
Predict_early_euth_df <- data.frame(Euth_Early = ifelse(meta_file[meta_file$day %in% 0,
                                                                  'Euthanized'] %in% 10,
                                                        'No','Yes'), 
                                    row.names = meta_file[meta_file$day %in% 0, 'Mouse_ID'])
Predict_early_euth_df <- merge(Predict_early_euth_df, rel_abund_stool, by='row.names')
Predict_early_euth_df <- Predict_early_euth_df[,!(colnames(Predict_early_euth_df) %in% "Row.names")]
rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df, 
                              importance = TRUE, ntree = n_trees)
# write(paste("binary", ncol(rel_abund_stool), rf_early_euth$err.rate[n_trees], "\t"), file="data/process/random_forest.data", append=T)
rf_early_euth
```

>>>>

```{r, message=FALSE}
pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

@@

for (i in 1:10){
  predict_all_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day' , 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
  predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
  rsq_df[i] <- RF$rsq[length(RF$rsq)]
  # sort the features by their effect on % increase in the standard error
  importance_df[ , i] <- importance(RF)[,1]
  colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
# fit the model back on the datar and tack it to the end of the counts file
  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
                              RF$predicted)
  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
  prediction_df <- rbind(prediction_df, random_forest_output)  
}

ggplot(data = prediction_df, aes(x=observed, y=predicted)) + geom_point() + 
        facet_wrap( ~ day, nrow = 2) + 
        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Day 0 Community',
        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')

rsq_df <- t(rsq_df)
rsq_df <- as.data.frame(cbind(rsq_df, day = c(1:10)))
ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
        x= 'Day', y = 'Variance Explained')

n_important <- 20
median_importance <- names(head(sort((apply(importance_df,1, median)), decreasing = T), n_important))
med_import_df <- melt(cbind(rownames(importance_df), importance_df)[median_importance, ])
colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
        geom_point() + scale_x_discrete(labels = c(1:10)) +
        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
        x= 'Day', y = '% Increase in MSE') + theme_bw()
        coord_flip()

#let's plot the top 12 features
#n_features <- 12
#importance_subset <- importance_sorted[1:n_features]
# let's get Rsq for the subset
#rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
#                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
#                           importance=TRUE, ntree=n_trees)
# write(paste("partial", n_features, rf_partial$rsq[n_trees], "\t"), file="data/process/random_forest.data", append=T)
```

@@

```{r}

CFU_euth <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
CFU_euth <- CFU_euth[ ,!(colnames(CFU_euth) %in% "Mouse_ID")]

RF_CFU_day_euth <- randomForest(log_cfu ~ ., data = CFU_euth[!is.na(CFU_euth$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
rsq_day_euth <- RF_CFU_day_euth$rsq[length(RF_CFU_day_euth$rsq)]
# sort the features by their effect on % increase in the standard error
importance_day_euth <- importance(RF_CFU_day_euth)[,1]
# fit the model back on the datar and tack it to the end of the counts file
rf_predict_cfu_output <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'day', 'log_cfu') ],
                              as.data.frame(predict(RF_CFU_day_euth, rel_abund_stool)), 
                              by='Mouse_ID', by.y='row.names')
colnames(rf_predict_cfu_output)[3:4] <- c('observed', 'predicted')

str(RF_CFU_day_euth)
TEST <- cbind(CFU_euth[!is.na(CFU_euth$log_cfu), 'log_cfu'], data.frame(RF_CFU_day_euth$predicted))
colnames(TEST) <- c('observed', 'predicted')
ggplot(data = TEST, aes(x=observed, y=predicted)) + geom_point()

ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
        labs(title = 'Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community',
        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw()
        ggtitle(expression(atop("Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community",
                                atop(paste('% Var Explained = ', rsq_day_euth), ""))))

top_important_OTU <- data.frame(head(sort(importance_day_euth, decreasing = T), n_important))
colnames(top_important_OTU) <- 'Importance'
top_important_OTU$OTU <- rownames(top_important_OTU)

ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
        geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
        x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()

@@

##Can we predict moribundity with the community structure at Day 0?
```{r}
##Fix to be categorical, and not regression
Predict_early_euth_df <- data.frame(Euth_Early = ifelse(meta_file[meta_file$day %in% 0,
                                                                  'Euthanized'] %in% 10,
                                                        'No','Yes'), 
                                    row.names = meta_file[meta_file$day %in% 0, 'Mouse_ID'])
Predict_early_euth_df <- merge(Predict_early_euth_df, rel_abund_stool, by='row.names')
Predict_early_euth_df <- Predict_early_euth_df[,!(colnames(Predict_early_euth_df) %in% "Row.names")]
rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df, 
                              importance = TRUE, ntree = n_trees)
# write(paste("binary", ncol(rel_abund_stool), rf_early_euth$err.rate[n_trees], "\t"), file="data/process/random_forest.data", append=T)
#varImpPlot(rf_early_euth)
#prediction_euth <- as.numeric(predict(rf_early_euth, Predict_early_euth_df[,-1]))
#ROC_euth <- roc(as.numeric(Predict_early_euth_df$Euth_Early), prediction_euth)
#plot(ROC_euth)
```
